\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-21 papers/jss/article.Rnw
Add citations for rjson and RJSONIO in the JSON section to make this comparable to the other formats described in the introduction which include citations to R packages.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@801 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Add a short sentence to define serialization early in the intro, addressing a todo that multiple people had mentioned.  Reference the C++ FAQ for lack of a better reference for now.
Revert the second to last paragraph of the introduction to my earlier
version that was instead moved to section 2.  Remove one of the more
techni

%% an abstract and keywords
\Abstract{
Modern data collection and analysis pipelines often involve
a sophisticated mix of applications written in general purpose and
specialized programming languages.  Protocol Buffers are a popular
method of serializing structured data between applications---while remaining
independent of programming languages or operating system.  The
\CRANpkg{RProtoBuf} package provides a complete interface between this
library and the R environment for statistical computing.
%TODO(ms) keep it less than 150 words.
}
\Keywords{r, protocol buffers, serialization, cross-platform}
\Plainkeywords{r, protocol buffers, serialization, cross-platform} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction} % TODO(DE) More sober: Friends don't let friends use CSV}
% NOTE(MS): I really do think we can use add back:
% \section{Introduction: Friends Don't Let Friends Use CSV}
% I didn't use proper Title Caps the first time around but really I
% think it makes the paper more readable to have a tl;dr intro title
% that is fun and engaging since this paper is still on the dry/boring
% side.
Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for the different phases of data analysis -- collection,
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?
% MS: yes perhaps somewhere near here we could define serialization
% and describe this.

Given these requirements, how do we safely and efficiently share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer systems?
% Reverted to my original above, because the replacement below puts me
% to sleep:
%Such systems require reliable and efficient exchange of intermediate
%results between the individual components, using formats that are
%independent of platform, language, operating system or architecture.
Programming
languages such as R, Julia, Java, and Python include built-in
support for serialization, but the default formats 
are usually language specific and thereby lock the user into a single
environment.

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character seperated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as seperator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexitiy at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.
%
A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects.

@@

%and \texttt{MessagePack} which both add binary support. However, these
%derivatives are not compatible with existing JSON software, and have not seen
%wide adoption.

%\paragraph*{Enter Protocol Buffers:}
In 2008, and following several years of internal use, Google released an open
source version of Protocol Buffers. It provides data  
interchange format that was designed and used for their internal infrastructure.
Google officially provides high-quality parsing libraries for \texttt{Java}, 
\texttt{C++} and \texttt{Python}, and community-developed open source implementations
are available for many other languages. 
Protocol Buffers take a quite different approach from many other popular formats.
They offer a unique combination of features, performance, and maturity that seems
particulary well suited for data-driven applications and numerical computing.
Protocol Buffers are a binary format that natively supports all common primitive types
found in modern programming languages. A key advantage is that numeric values
are serialized exactly the same way as they are stored in memory. There is
no loss of precision, no overhead, and parsing messages is very efficient: the system can 
simply copy bytes to memory without any further processing. 
But the most powerful feature of Protocol Buffers is that it decouples the content
from the structure using a schema, very similar to a database. This further increases
performance by eliminating redundancy, while at the same time providing foundations
for defining an \emph{Interface Description Language}, or \emph{IDL}.
Many sources compare data serialization formats and show Protocol Buffers compare 
very favorably to the alternatives; see \citet{Sumaray:2012:CDS:2184751.2184810} 
for one such comparison.
% TODO(DE): Mention "future proof" forward compatibility of schemata

%  The schema can be used to
%generate model classes for statically-typed programming languages such
%as C++ and Java, or can be used with reflection for dynamically-typed
%programming languages.  Since the schema is provided separately from
%the encoded data, the data can be efficiently encoded to minimize
%storage costs of the stored data when compared with simple
%``schema-less'' binary interchange formats.
%Many sources compare data serialization formats and show Protocol
%Buffers very

@@

% TODO(mstokely): If we go to JSS, include a larger paragraph here
% referencing each numbered section.  I don't like these generally,
% but its useful for this paper I think because we have a boring bit
% in the middle (full class/method details) and interesting
% applications at the end.

This paper describes an R interface to Protocol Buffer, 
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

>>>>

%% an abstract and keywords
\Abstract{
Modern data collection and analysis pipelines often involve
a sophisticated mix of applications written in general purpose and
specialized programming languages.  Protocol Buffers are a popular
method of serializing structured data between applications---while remaining
independent of programming languages or operating system.  The
\CRANpkg{RProtoBuf} package provides a complete interface between this
library and the R environment for statistical computing.
%TODO(ms) keep it less than 150 words.
% Maybe add Jeroen's sentence:
% They offer a unique combination of features, performance, and maturity that seems
% particulary well suited for data-driven applications and numerical
% computing.
}
\Keywords{r, protocol buffers, serialization, cross-platform}
\Plainkeywords{r, protocol buffers, serialization, cross-platform} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction} % TODO(DE) More sober: Friends don't let friends use CSV}
% NOTE(MS): I really do think we can use add back:
% \section{Introduction: Friends Don't Let Friends Use CSV}
% I didn't use proper Title Caps the first time around but really I
% think it makes the paper more readable to have a tl;dr intro title
% that is fun and engaging since this paper is still on the dry/boring
% side.
Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for the different phases of data analysis -- collection,
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?

Given these requirements, how do we safely and efficiently share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer systems?
In computer programming, \emph{serialization} is the process of
translating data structures, variables, and session state into a
format that can be stored or transmitted and then reconstructed in the
original form later \citep{clinec++}.
% Reverted to my original above, because the replacement below puts me
% to sleep:
%Such systems require reliable and efficient exchange of intermediate
%results between the individual components, using formats that are
%independent of platform, language, operating system or architecture.
Programming
languages such as R, Julia, Java, and Python include built-in
support for serialization, but the default formats 
are usually language specific and thereby lock the user into a single
environment.

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character seperated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as seperator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexitiy at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.
%
A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R
objects \citep{rjson,RJSONIO}.

@@

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro
provide a compact well-documented schema for cross-language data
structures and efficient binary interchange formats.  Since the schema
is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs when
compared with simple ``schema-less'' binary interchange formats.
Many sources compare data serialization formats
and show Protocol Buffers compare very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.

% Too technical, move to section 2.
% The schema can be used to generate model classes for statically-typed programming languages
%such as C++ and Java, or can be used with reflection for dynamically-typed programming
%languages.

% TODO(mstokely): Will need to define reflection if we use it here.
% Maybe in the next section since its not as key as 'serialization'
% which we already defined.

%\paragraph*{Enter Protocol Buffers:}

% In 2008, and following several years of internal use, Google released an open
% source version of Protocol Buffers. It provides data  
% interchange format that was designed and used for their internal infrastructure.
% Google officially provides high-quality parsing libraries for \texttt{Java}, 
% \texttt{C++} and \texttt{Python}, and community-developed open source implementations
% are available for many other languages. 
% Protocol Buffers take a quite different approach from many other popular formats.

% TODO(mstokely): Good sentence from Jeroen, add it here or sec 2.
% They offer a unique combination of features, performance, and maturity that seems
% particulary well suited for data-driven applications and numerical
% computing.

% TODO(DE): Mention "future proof" forward compatibility of schemata

@@

% NOTE(mstokely): I don't like these roadmap paragraphs in general,
% but it seems ueful here because we have a boring bit in the middle
% (full class/method details) and interesting applications at the end.

This paper describes an R interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.
